Generalized spaces for constructive algebra

### 2020-04-08 paper-herrsching2019/main.tex
Fixed a couple of typos (thanks to aspell)

However, these reduction techniques typically require \emph{transfinite
methods}. The displayed proof appeals to the transfinite four times:
\begin{enumerate}
\item The proof proceeds by contradiction, hence requires the law of excluded
middle, the strongest of the omniscience principles.
\item The proof then requires a minimal prime ideal. Such an ideal can be
obtained in two steps: First, the Boolean Prime Ideal
Theorem~\BPIT is invoked to obtain some prime ideal. Then
Zorn's lemma fabricates a minimal one.\footnote{Standard textbooks prove the statement ``any nontrivial
ring has a prime ideal'' by using Zorn's lemma, which
in the presence of the law of excluded middle is equivalent to the full axiom
of choice; however, the statement is actually equivalent to the weaker Boolean
Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
The combined statement ``any ring has a minimal prime ideal'' is equivalent to
the full axiom of choice~\cite{savin:minimal-prime-ideals}.}
%FUTURE:
%This observation might be folklore; we learned the
%following argument from Will Savin~[reference]: Let~$(M_i)_{i \in I}$ be a
%disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
%its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
%that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
%preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
%minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
%there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
%index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
%none, then pick an element~$a \in M_i$ ...}
\item Finally, the proof exploits that the stalks of reduced rings at minimal
prime ideals are fields. This requires two further invokations
of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
is its nilradical. Since~$A_\ppp$ is reduced, the nilradical is the zero ideal.
Since the prime ideals of~$A_\ppp$ are in bijection with those prime ideals
of~$A$ which are contained in~$\ppp$, the ring~$A_\ppp$ has exactly one prime
ideal. By~\BPIT, any noninvertible element of~$A_\ppp$ is contained in some
prime ideal, hence in the zero ideal. Thus~$A_\ppp$ is a field.}
\end{enumerate}

@@

In some cases, these new techniques go even beyond their classical role models.
For instance, while a module need not be free ``on a dense open'' if all its
stalks at minimal prime ideals are, a module~$M$ is free in that sense if its
mirror image~$M^\sim$ is \notnot free. This observation is the basis for a
new proof of Grothendieck's generic freeness lemma, reviewed in
Section~\ref{sect:example-applications}, which is shorter and arguably more
perspicious than all previously known classical proofs.

@@

\noindent{\scriptsize\emph{Note.} In the presense of the countable axiom of choice, any
Cauchy process can be refined to a Cauchy sequence. In its absence, however, the
familiar completion construction using equivalence classes of Cauchy sequences
fails to yield a Cauchy-complete space. In contrast, any Cauchy sequence and
more generally any Cauchy process converges in the space of equivalence classes
of Cauchy processes.\par}
\end{exercise}

@@

The localic open unit interval is the classifying locale of the
same theory but with the additional axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
\item The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
classifying locale of the theory of surjections~$\NN \twoheadrightarrow \RR$.
\item The \emph{spectrum} of a ring~$A$ is the classifying locale of the
theory of prime filters of~$A$. The topological space of points of this locale
is the Zariski spectrum as familiar from algebraic
geometry.\footnote{Classicaly, the spectrum is usually defined as the
topological space of prime ideals instead of prime filters. However,
classically there is a canonical one-to-one correspondence between prime ideals
and prime filters, hence one could just as well use filters instead of ideals
in the definition. The classifying locale of prime ideals of~$A$ also exists,
and has the prime ideals of~$A$ as its points, but the induced topology on its
topological space of points is the \emph{flat topology} or \emph{co-Zariski
topology}~\cite{tarizadeh:flat} instead of the ordinary
Zariski topology~\cite[Proposition~4.5]{johnstone:rings-fields-and-spectra}.}
\item The \emph{localic Galois group} of a field extension~$L|k$ is the
classifying locale of the theory of ring automorphisms~$L \to L$ which fix~$k$.
\item The \emph{empty locale} is the classifying locale of the inconsistent theory.
\item The \emph{one-point locale} is the classifying locale of the empty theory.
%\item The \emph{discrete locale on a set~$M$} is the classifying locale of the
%theory which has one nullary relation symbol~$\varphi_x$ for each element~$x
%\in M$, the axiom~$\top \vdash \bigvee_{x \in M} \varphi_x$ and for each~$x,y
%\in M$ the axiom~$\varphi_x \wedge \varphi_y \vdash \bigvee\{ \top \,|\, x = y
%\}$.
\end{enumerate}

@@

In general, the resulting proofs will be longer and less perspicious
than the original proofs, in particular if the original proofs employed nested
implications (such as double negations). In these cases, there is a certain
``exotic flavor'' to the unwound proofs, as with tricks with continuations in
computer science. We display some such proofs which are still readable in
Section~\ref{sect:example-applications}.
\end{remark}

@@

We can also cast the contruction in logical language. This change in perspective makes it evident
that abelianization is just a special case of a more general procedure.
The free abelian group over~$G$ can be obtained as the \emph{term model} of the
theory of abelian groups extended by a constant~$x_g$ for each element~$g \in
G$ and the axioms
\begin{align*}
  \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
  \top &\vdash x_{g^{-1}} = x_g^{-1} & \text{(for each~$g \in G$)} \\
  \top &\vdash x_e = e
\end{align*}
That is, $G^\ab$ contains exactly those elements and exactly those
identifications between elements such that~$G^\ab$ is an abelian group and such
that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
G^\ab$), and nothing more.

In exactly the same manner, we can construct the free group over a monoid, the
free~$\ZZ/(2)$-algebra over a ring, the free ring over a set or the free
Heyting algebra over a partially ordered set. This is all well-understood; a
particularly perspicious account in the general setting of partial Horn
theories is due to Erik Palmgren and Steve Vickers~\cite[Section~5]{palmgren-vickers:partial-horn}.

@@

In exactly the same fashion, the dual statement about surjective matricen can
be proven.

@@

\begin{remark}Not of pragmatic, but of logical interest is that the proof can
also be recast as follows. Instead of reducing to the factor ring~$A/\ann(x)$,
we continue considering~$A$. Working under the sheaf sematics, we then
introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow a = 0)$.
We may still use the law of excluded middle for this new negation and hence
exploit that any element~$x$ of~$A^\sim$ is invertible or ``not'' invertible. By
Exercise~\ref{ex:gen-field-property}, this amounts to saying that some power
of~$x$ annihilates~$a$. The remainder of the proof carries over with minor
modifications.
\end{remark}

@@

It is on purpose that the (sheaf-theoretic version of) our proof contains case
distinctions, to closely mimick the proof in the field case. It is also
possible to do without, as already our unrolled proof demonstrates. Thierry
Coquand and Claude Quitté gave a beautiful proof without case distinctions an
application of a local--global
principle~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions};
if desired, their proof can also be cast in the sheaf-theoretic framework
(Exercise~\ref{ex:local-global}).

@@

{\scriptsize\emph{Hint.} For a truth value~$\varphi$, consider the ideal~$\{ x
\in \ZZ \,|\, x = 0 \vee \varphi \}$.\par}
\item Give an example of a space~$X$ and a subsheaf~$F$ of the constant
sheaf~$\underline{\ZZ}$ such that, under the sheaf semantics, the subsheaf~$F$
is an ideal, and such that it is not the case that, again under the sheaf
semantics, the ideal~$F$ is finitely generated.
\item Give a proof that~$\ZZ$ is anonymously Noetherian which, unlike the proof
given in Example~\ref{ex:z-noeth}, uses the anonymous version of the least
numbr principle of Exercise~\ref{ex:least-number-principle}.
\end{enumerate}
\end{exercise}

>>>>

However, these reduction techniques typically require \emph{transfinite
methods}. The displayed proof appeals to the transfinite four times:
\begin{enumerate}
\item The proof proceeds by contradiction, hence requires the law of excluded
middle, the strongest of the omniscience principles.
\item The proof then requires a minimal prime ideal. Such an ideal can be
obtained in two steps: First, the Boolean Prime Ideal
Theorem~\BPIT is invoked to obtain some prime ideal. Then
Zorn's lemma fabricates a minimal one.\footnote{Standard textbooks prove the statement ``any nontrivial
ring has a prime ideal'' by using Zorn's lemma, which
in the presence of the law of excluded middle is equivalent to the full axiom
of choice; however, the statement is actually equivalent to the weaker Boolean
Prime Ideal Theorem~\cite{scott:bpit,banaschewski-harting:lattice-aspects}.
The combined statement ``any ring has a minimal prime ideal'' is equivalent to
the full axiom of choice~\cite{savin:minimal-prime-ideals}.}
%FUTURE:
%This observation might be folklore; we learned the
%following argument from Will Savin~[reference]: Let~$(M_i)_{i \in I}$ be a
%disjoint family of inhabited sets. We consider the free ring~$\ZZ[\bigcup_i M_i]$ and
%its ideal~$I$ generated by~$\{ ab \,|\, i \in I, a,b \in M_i, a \neq b \}$. Assume
%that the factor ring~$\ZZ[\bigcup_i M_i]/I$ contains a minimal prime ideal. Its
%preimage~$\ppp \subseteq \ZZ[\bigcup_i M_i]$ is then a prime ideal which is
%minimal among all prime ideals containing~$I$. By primality, for any index~$i$,
%there is at most one element of~$M_i$ not in~$\ppp$. By minimality, for each
%index~$i$, there is exactly one element of~$M_i$ not in~$\ppp$: If there were
%none, then pick an element~$a \in M_i$ ...}
\item Finally, the proof exploits that the stalks of reduced rings at minimal
prime ideals are fields. This requires two further invocations
of~\BPIT.\footnote{By~\BPIT, the intersection of all prime ideals of~$A_\ppp$
is its nilradical. Since~$A_\ppp$ is reduced, the nilradical is the zero ideal.
Since the prime ideals of~$A_\ppp$ are in bijection with those prime ideals
of~$A$ which are contained in~$\ppp$, the ring~$A_\ppp$ has exactly one prime
ideal. By~\BPIT, any noninvertible element of~$A_\ppp$ is contained in some
prime ideal, hence in the zero ideal. Thus~$A_\ppp$ is a field.}
\end{enumerate}

@@

In some cases, these new techniques go even beyond their classical role models.
For instance, while a module need not be free ``on a dense open'' if all its
stalks at minimal prime ideals are, a module~$M$ is free in that sense if its
mirror image~$M^\sim$ is \notnot free. This observation is the basis for a
new proof of Grothendieck's generic freeness lemma, reviewed in
Section~\ref{sect:example-applications}, which is shorter and arguably more
perspicuous than all previously known classical proofs.

@@

\noindent{\scriptsize\emph{Note.} In the presence of the countable axiom of choice, any
Cauchy process can be refined to a Cauchy sequence. In its absence, however, the
familiar completion construction using equivalence classes of Cauchy sequences
fails to yield a Cauchy-complete space. In contrast, any Cauchy sequence and
more generally any Cauchy process converges in the space of equivalence classes
of Cauchy processes.\par}
\end{exercise}

@@

The localic open unit interval is the classifying locale of the
same theory but with the additional axiom~$\top \vdash \alpha_0 \wedge \beta_1$.
\item The \emph{locale of surjections~$\NN \twoheadrightarrow \RR$} is the
classifying locale of the theory of surjections~$\NN \twoheadrightarrow \RR$.
\item The \emph{spectrum} of a ring~$A$ is the classifying locale of the
theory of prime filters of~$A$. The topological space of points of this locale
is the Zariski spectrum as familiar from algebraic
geometry.\footnote{Classically, the spectrum is usually defined as the
topological space of prime ideals instead of prime filters. However,
classically there is a canonical one-to-one correspondence between prime ideals
and prime filters, hence one could just as well use filters instead of ideals
in the definition. The classifying locale of prime ideals of~$A$ also exists,
and has the prime ideals of~$A$ as its points, but the induced topology on its
topological space of points is the \emph{flat topology} or \emph{co-Zariski
topology}~\cite{tarizadeh:flat} instead of the ordinary
Zariski topology~\cite[Proposition~4.5]{johnstone:rings-fields-and-spectra}.}
\item The \emph{localic Galois group} of a field extension~$L|k$ is the
classifying locale of the theory of ring automorphisms~$L \to L$ which fix~$k$.
\item The \emph{empty locale} is the classifying locale of the inconsistent theory.
\item The \emph{one-point locale} is the classifying locale of the empty theory.
%\item The \emph{discrete locale on a set~$M$} is the classifying locale of the
%theory which has one nullary relation symbol~$\varphi_x$ for each element~$x
%\in M$, the axiom~$\top \vdash \bigvee_{x \in M} \varphi_x$ and for each~$x,y
%\in M$ the axiom~$\varphi_x \wedge \varphi_y \vdash \bigvee\{ \top \,|\, x = y
%\}$.
\end{enumerate}

@@

In general, the resulting proofs will be longer and less perspicuous
than the original proofs, in particular if the original proofs employed nested
implications (such as double negations). In these cases, there is a certain
``exotic flavor'' to the unwound proofs, as with tricks with continuations in
computer science. We display some such proofs which are still readable in
Section~\ref{sect:example-applications}.
\end{remark}

@@

We can also cast the construction in logical language. This change in perspective makes it evident
that abelianization is just a special case of a more general procedure.
The free abelian group over~$G$ can be obtained as the \emph{term model} of the
theory of abelian groups extended by a constant~$x_g$ for each element~$g \in
G$ and the axioms
\begin{align*}
  \top &\vdash x_{g \circ h} = x_g \circ x_h & \text{(for each~$g,h \in G$)} \\
  \top &\vdash x_{g^{-1}} = x_g^{-1} & \text{(for each~$g \in G$)} \\
  \top &\vdash x_e = e
\end{align*}
That is, $G^\ab$ contains exactly those elements and exactly those
identifications between elements such that~$G^\ab$ is an abelian group and such
that~$G$ can be interpreted in it (that is, that there is a homomorphism~$G \to
G^\ab$), and nothing more.

In exactly the same manner, we can construct the free group over a monoid, the
free~$\ZZ/(2)$-algebra over a ring, the free ring over a set or the free
Heyting algebra over a partially ordered set. This is all well-understood; a
particularly perspicuous account in the general setting of partial Horn
theories is due to Erik Palmgren and Steve Vickers~\cite[Section~5]{palmgren-vickers:partial-horn}.

@@

In exactly the same fashion, the dual statement about surjective matrices can
be proven.

@@

\begin{remark}Not of pragmatic, but of logical interest is that the proof can
also be recast as follows. Instead of reducing to the factor ring~$A/\ann(x)$,
we continue considering~$A$. Working under the sheaf semantics, we then
introduce a variant of the usual negation~$\neg\varphi \equiv (\varphi
\Rightarrow \bot)$, namely~$\negg\varphi \equiv (\varphi \Rightarrow a = 0)$.
We may still use the law of excluded middle for this new negation and hence
exploit that any element~$x$ of~$A^\sim$ is invertible or ``not'' invertible. By
Exercise~\ref{ex:gen-field-property}, this amounts to saying that some power
of~$x$ annihilates~$a$. The remainder of the proof carries over with minor
modifications.
\end{remark}

@@

It is on purpose that the (sheaf-theoretic version of) our proof contains case
distinctions, to closely mimic the proof in the field case. It is also
possible to do without, as already our unrolled proof demonstrates. Thierry
Coquand and Claude Quitté gave a beautiful proof without case distinctions an
application of a local--global
principle~\cite[Theorem~2.4]{coquand-quitte:constructive-finite-free-resolutions};
if desired, their proof can also be cast in the sheaf-theoretic framework
(Exercise~\ref{ex:local-global}).

@@

{\scriptsize\emph{Hint.} For a truth value~$\varphi$, consider the ideal~$\{ x
\in \ZZ \,|\, x = 0 \vee \varphi \}$.\par}
\item Give an example of a space~$X$ and a subsheaf~$F$ of the constant
sheaf~$\underline{\ZZ}$ such that, under the sheaf semantics, the subsheaf~$F$
is an ideal, and such that it is not the case that, again under the sheaf
semantics, the ideal~$F$ is finitely generated.
\item Give a proof that~$\ZZ$ is anonymously Noetherian which, unlike the proof
given in Example~\ref{ex:z-noeth}, uses the anonymous version of the least
number principle of Exercise~\ref{ex:least-number-principle}.
\end{enumerate}
\end{exercise}
